Community context determines competition/facilitation trade-offs in pollination systems

### 2019-04-15 paper/manuscript.Rmd
Add tentative title
Add some of Hugo's feedback to discussion

---
title: "The pollination trade-off"
author: 
  - "E. Fernando Cagua^1^ ([efc29@uclive.ac.nz](mailto:efc29@uclive.ac.nz))"
  - "Hugo J. Marrero^2^ ([hugomarrero@gmail.com](hugomarrero@gmail.com))"
  - "Jason M. Tylianakis^1^ ([jason.tylianakis@canterbury.ac.nz](jason.tylianakis@canterbury.ac.nz))"
  - "Daniel B. Stouffer^1^ ([daniel.stouffer@canterbury.ac.nz](daniel.stouffer@canterbury.ac.nz))"
bibliography: phd-bibliography/pollen-competition.bib
csl: ecology-letters.csl
output: #word_document
  bookdown::pdf_document2:
    # base_format: rticles::peerj_article
    keep_tex: yes
    number_sections: false
toc: false
fontsize: 11pt
classoption: a4paper
geometry: 
  - textwidth=33em
  - textheight=48\baselineskip
header-includes:
  - \usepackage{booktabs}
  - \usepackage{setspace}
  - \usepackage{lineno}
  - \usepackage{xr}
  - \usepackage[utf8]{inputenc}
  - \newcommand{\R}[1]{\label{#1}\linelabel{#1}}
  - \newcommand{\lr}[1]{page~\pageref{#1}, line~\lineref{#1}}
  - \externaldocument[S-]{supp-info}
  - \usepackage[export]{adjustbox}
---

@@

All else being equal, our results suggest that community context plays a central role in determining net cost or benefit of sharing pollinators.
First, we found that the costs of pollinator sharing are pervasive. 
In particular, we found that a considerable proportion of species experiences overall neutral or competitive effects from animal-mediated pollination. 
Second, we found that multiple ecological factors can modulate the intensity of this competition; however, conspecific and heterospecific pollen deposition are tightly coupled and therefore these factors are likely to involve a trade-off between the quantity and quality of pollination. 
Third, we found that the way these factors influence competition is strongly shaped by the community context, this is, if collectively these factors are used to define a niche of competition for pollination, the same plant species can occupy dramatically different niches depending on the community it belongs to.

Both of the tests we used to identify the effect of animal-mediated pollination suggested the same outcome: that net facilitation is only experienced by a fraction of species in a community. 
Instead, we observed a negative relationships between conspecific and heterospecific in a considerable proportion of the analysed species. 
This relationship could arise, for example, when another, more dominant species, monopolizes a limited number of pollinators [@mitchell_new_2009; @lopezaraizamikel_impact_2007], when heterospecific pollen deposited in the stigma phisically interferes with deposition of conspecific pollen [@runquist_pollinator-mediated_2012], or when conspecific pollen is lost to heterospecific stigmas [@mitchell_new_2009; @muchhala_interspecific_2012]. 
Similarly, not all species had more pollen deposited when open to animal pollination compared to when they were closed. 
Although it can be important, we were unable to test the effect of pollen viability or the extent to which self- vs. outcross-pollen may differ in their value for fertilization (TODO: ref needed). 
However, it is reasonable to expect that the difference between these two may be generally relevant. 
Particularly so for species that are not self-fertile or those in which self-fertilization is rarely effective due to a temporary separation in the maturation of the sexes (dichogamy) (TODO: ref needed). 
These findings agree with a recent study showing a wide range of outcomes in a multi-species context [@tur_evaluating_2016]. 
However, the factors explaining why some species experience more competition than others were not understood. 
Here we show that the prevalence of neutral and competitive effects of animal-mediated pollination can be at least partially explained by the trade-offs that minimising competition can involve.

These trade-offs were most evident in precisely the factors that most strongly modulate pollen deposition. 
Specifically, we found that the factors related to the relative dominance of the species (in terms of pollen and visits, or abundance) also tended to be the most important predictors of pollen deposition. 
At the same time, receiving high visitation increases the absolute levels of conspecific pollen deposition but increases heterospecific deposition to a greater extent---even when the visitors are likely to carry a high proportion of conspecific pollen. 
Similarly, being abundant reduces the amount of heterospecific pollen deposited but at the expense of a noticeable reduction in the amount of conspecific pollen. 
Multiple studies recognise the importance of visitation and abundance to competition for pollination [@feldman_when_2004; @munoz_presence_2008; @morales_interspecific_2008]. 
Our results strongly corroborate the importance of these two factors but also suggest that [because visitation, pollen production and abundance are correlated; @sargent_role_2006], plants must reach a fine balance between increasing conspecific pollen and reduce heterospecific pollen deposition to minimise competition for pollination.

Further evidence of the intricacies of balancing costs and benefits can be found when we examine the other two factors analysed in our models. 
We show that being a specialist and sharing no pollinators might be preferable in the short term (as it is generally detrimental to pollen deposition). 
However, over long periods of time, there could be a risk associated with a specialist plant having no pollinators. 
Hence, is likely that to ensure long term survival, plants also need to balance this risk with the costs of sharing pollinators. 
One possible solution is to share pollinators *and* have original traits---as we show that trait originality is generally beneficial to pollen deposition and it’s commonly thought that species that are further from others in trait space benefit from reduced competition. 
Yet again, there are two possible caveats to this strategy that highlight the interrelatedness of the ecological factors. 
First, in a mutualism context, it is also possible that trait originality could come at the cost of being less ‘apparent’ to pollinators. 
Second, the usually negative, relationship between originality and generalism has been shown depend on the plant's abundance [@coux_linking_2016], with generalist species being able to also have original traits only when they are abundant enough to provide a valuable reward to make it worth to pollinators.

How precisely plants balance the costs and benefits in the short and the long term likely depends on the species sensitivity to different risks but also, as we show here, on the community it belongs to. 
For instance, a plant species that experience a large reduction in seed set due to heterospecific pollen deposition might be better off by sharing few pollinators, while sharing pollinators makes more sense for those mainly affected by low levels of conspecific pollen deposition [@campbell_mechanism_1985].
Nevertheless, the observed diversity of the facilitation/competition outcomes across communities and the relevance of random effects in our models suggests that the community context has a very important effect on the way plant species respond to competition for pollination beyond species identity. 
Although there are exceptions, some plant families (Asteraceae for example) are often generalists in their communities while others (Orchidaceae) are known to be more commonly specialised. 
While it is possible, from an evolutionary perspective, that selection for a particular strategy might say something about the community in which a species has typically inhabited during its evolutionary history, we show that most plant species are still flexible enough to adopt markedly different niches in different communities. 
From a more applied perspective, flowering plants are sometimes introduced to attract pollinators on other nearby plants. 
Our results highlight the difficulties involved in predicting whether the introduced plant species will facilitate or compete with neighbours. 
This is so because the outcome might depend not only on the particular niche each species occupies in a particular community but also on other factors that we were unable to measure [e.g. pollinator behaviour and densities or spatial context; @ye_competition_2014; @flanagan_effects_2011; @cariveau_spatially_2009].
However, our results also suggest that introduced species are more likely to be beneficial if they have different traits to local plants. 
<!-- cite melissa's paper if available before publication -->

@@

>>>>

---
title: "Community context determines competition/facilitation trade-offs in pollination systems"
author: 
  - "E. Fernando Cagua^1^ ([efc29@uclive.ac.nz](mailto:efc29@uclive.ac.nz))"
  - "Hugo J. Marrero^2^ ([hugomarrero@gmail.com](hugomarrero@gmail.com))"
  - "Jason M. Tylianakis^1^ ([jason.tylianakis@canterbury.ac.nz](jason.tylianakis@canterbury.ac.nz))"
  - "Daniel B. Stouffer^1^ ([daniel.stouffer@canterbury.ac.nz](daniel.stouffer@canterbury.ac.nz))"
bibliography: phd-bibliography/pollen-competition.bib
csl: ecology-letters.csl
output: #word_document
  bookdown::pdf_document2:
    # base_format: rticles::peerj_article
    keep_tex: yes
    number_sections: false
toc: false
fontsize: 11pt
classoption: a4paper
geometry: 
  - textwidth=33em
  - textheight=48\baselineskip
header-includes:
  - \usepackage{booktabs}
  - \usepackage{setspace}
  - \usepackage{lineno}
  - \usepackage{xr}
  - \usepackage[utf8]{inputenc}
  - \newcommand{\R}[1]{\label{#1}\linelabel{#1}}
  - \newcommand{\lr}[1]{page~\pageref{#1}, line~\lineref{#1}}
  - \externaldocument[S-]{supp-info}
  - \usepackage[export]{adjustbox}
---

@@

Our results suggest that community context plays a central role in determining net cost or benefit of sharing pollinators.
First, we found that the costs of pollinator sharing are pervasive. 
In particular, we found that a considerable proportion of species experiences overall neutral or competitive effects from animal-mediated pollination. 
Second, we found that multiple ecological factors can modulate the intensity of this competition; however, conspecific and heterospecific pollen deposition are tightly coupled and therefore these factors are likely to involve a trade-off between the quantity and quality of pollination. 
Third, we found that the way these factors influence competition is strongly shaped by the community context, this is, if collectively these factors are used to define a niche of competition for pollination, the same plant species can occupy dramatically different niches depending on the community it belongs to.

Both of the criteria we used to identify the effect of animal-mediated pollination suggested the same outcome: that neither facilitation or competition are dominant in pollination communities.
A previous study showed that in diverse pollination communities, sharing pollinators does not necessarily translate to net competition [@tur_evaluating_2016]. 
The overall extent of facilitation/competition of pollination communities can depend on factors like the adversity of the environment [@tur_evaluating_2016; @callaway_positive_2002] or how disturbances affect pollinator populations [@stavert_exotic_2017]
However, why some species experience more competition than others was not understood. 
Here we show that the coexistence of facilitative and competitive effects of animal-mediated pollination can be at least partially explained by the trade-offs plants have to undergo when maximising the deposition of conspecific pollen and simultaneously minimising that of heterospecific pollen both in the short and the long term.

In the short term, being a specialist and sharing no pollinators might reduce competition [@muchhala_competition_2010] and hence be preferable, as we show that sharing pollinators reduces conspecific pollen deposition both in absoluate terms and relative to heterospecific pollen. 
However, over long periods of time, there could be a risk associated with a specialist plant having no pollinators. 
Hence, is likely that to ensure long term survival, plants also need to balance this risk with the costs of sharing pollinators. 
One possible solution is to share pollinators *and* have original traits---as we show that trait originality is generally beneficial to pollen deposition and it’s commonly thought that species that are further from others in trait space benefit from reduced competition. 
Yet, there are two possible caveats to this strategy that highlight the interrelatedness of the ecological factors. 
First, in a mutualism context, it is also possible that trait originality could come at the cost of being less ‘apparent’ to pollinators. 
Second, the usually negative, relationship between originality and generalism has been shown depend on the plant's abundance [@coux_linking_2016], with generalist species being able to also have original traits only when they are abundant enough to provide a valuable reward to make it worth to pollinators.

Visit effectiveness (high pollen and visits) and abundance, which were the most important predictors of pollen deposition, involved an even more explicit trade-off between conspecific and heterospecific pollen. 
On the one hand, receiving high visitation increases the absolute levels of conspecific pollen deposition but increases heterospecific deposition to a greater extent---even when the visitors are likely to carry a high proportion of conspecific pollen. 
On the other, being abundant reduces the amount of heterospecific pollen deposited but at the expense of a noticeable reduction in the amount of conspecific pollen. 
Our results corroborate the importance that two-species studies have ascribed to visitation and abundance [@feldman_when_2004; @munoz_presence_2008; @morales_interspecific_2008]. 
But importantly, they also suggest that [because visitation, pollen production and abundance are usually correlated; @sargent_role_2006] balancing the pros and cons of sharing pollinators at the community level is not trivial and might even be partially responsible for the diversity of plant-pollinator communities [@benadi_frequency_2018].

How precisely plants balance the costs and benefits in the short and the long term likely depends on the species sensitivity to different risks but also, as we show here, on the community it belongs to. 
<!-- The mechanisms through which sharing pollinators may induce net competition are relatively well understood.  -->
<!-- It can occurr for example, when a more dominant species, monopolizes a limited number of pollinators [@mitchell_new_2009; @lopezaraizamikel_impact_2007], when heterospecific pollen deposited in the stigma phisically interferes with deposition of conspecific pollen [@runquist_pollinator-mediated_2012], or when conspecific pollen is lost to heterospecific stigmas [@mitchell_new_2009; @muchhala_interspecific_2012].  -->
<!-- Similarly, not all species had more pollen deposited when open to animal pollination compared to when they were closed.  -->
<!-- Although it can be important, we were unable to test the effect of pollen viability or the extent to which self- vs. outcross-pollen may differ in their value for fertilization (TODO: ref needed).  -->
<!-- However, it is reasonable to expect that the difference between these two may be generally relevant.  -->
<!-- Particularly so for species that are not self-fertile or those in which self-fertilization is rarely effective due to a temporary separation in the maturation of the sexes (dichogamy) (TODO: ref needed).  -->
For instance, a plant species that experience a large reduction in seed set due to heterospecific pollen deposition might be better off by sharing few pollinators, while sharing pollinators makes more sense for those mainly affected by low levels of conspecific pollen deposition [@campbell_mechanism_1985].
Nevertheless, the observed diversity of the facilitation/competition outcomes across communities and the relevance of random effects in our models suggests that the community context has a very important effect on the way plant species respond to competition for pollination beyond species identity. 
Although there are exceptions, some plant families (Asteraceae for example) are often generalists in their communities while others (Orchidaceae) are known to be more commonly specialised. 
While it is possible, from an evolutionary perspective, that selection for a particular strategy might say something about the community in which a species has typically inhabited during its evolutionary history, we show that most plant species are still flexible enough to adopt markedly different niches in different communities. 
From a more applied perspective, flowering plants are sometimes introduced to attract pollinators on other nearby plants. 
Our results highlight the difficulties involved in predicting whether the introduced plant species will facilitate or compete with neighbours. 
This is so because the outcome might depend not only on the particular niche each species occupies in a particular community but also on other factors that we were unable to measure [e.g. pollinator behaviour and densities or spatial context; @ye_competition_2014; @flanagan_effects_2011; @cariveau_spatially_2009].
However, our results also suggest that introduced species are more likely to be beneficial if they have different traits to local plants. 
<!-- cite melissa's paper if available before publication -->

@@

Extra potential refs:
https://besjournals.onlinelibrary.wiley.com/doi/pdf/10.1111/1365-2745.12256
https://academic.oup.com/aob/article/118/2/249/1741474
https://www.jstor.org/stable/3547899?seq=1#metadata_info_tab_contents
https://onlinelibrary.wiley.com/doi/full/10.1111/ele.12342
